A note on the instability and degeneracy of deep learning models

### 2016-10-22 writing/note.Rmd
2nd edits from Nordman

---
title: "A note on the instability and degeneracy of deep learning models"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "October, 2016"
linestretch: 1.5
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
---

@@

Essentially, in the process of developing a model to handle increasing numbers of variable (i.e., $N \rightarrow \infty$), we want (\ref{eqn:elpr}) to stay bounded, meaning that the largest probability under this model should maintain a magnitude within the smallest probability allowed under the model. This leads to the following definition:

\begin{defn}[unstable DEF]
A DEF model formulation is \emph{unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that 
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.

Note that this definition of *unstable* is a generalization of that in @schweinberger2011instability allowing nonexponential family models and an increasing number of parameters (while this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there, see Section \ref{discrete-exponential-family-models} for details).

Unstable DEF models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta_N(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
$$
the biggest log-probability ratio for a one component change in data outcomes. We then have the following result.

\begin{prop}
\label{prop:instab}
If $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$, then $\Delta_N(\boldsymbol \theta_N) > C$.
\end{prop}

In other words, if the quantity (\ref{eqn:elpr}) is too large, then the DEF model will exhibit large probability shifts for very small changes in the data configuration (i.e. instability mentioned earlier).

**Proof of Proposition \ref{prop:instab}.** We prove the contrapositive. Suppose that $\Delta_N(\boldsymbol \theta_N) \le C$ holds for some $C > 0$. Under the DEF model, $P_{\boldsymbol \theta_V}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$. Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$. Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e. $\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$. Then
\begin{align*}
\log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \\
& \le k \Delta_N(\boldsymbol \theta_N) \le NC
\end{align*}
using $k \le N$ and $\Delta_N(\boldsymbol \theta_N) \le C$.$\Box$

Additionally, unstable RBM models are connected to degenerate models, where *degeneracy* implies placing all probability on a small pieace of the sample space. Define a model set 
$$
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{(x_1, \dots, x_N) \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > (1-\epsilon)\max\limits_{(x_1^*, \dots, x_N^*)}P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*) + \epsilon\min\limits_{(x_1^*, \dots, x_N^*)}P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*) \right\}
$$
of possible outcomes, for a given $0 < \epsilon < 1$.

@@

In other words, as a consequence of unstable models, all probability will stack up on mode sets or potentially those few outcomes with the highest probability.

**Proof of Proposition \ref{prop:degen}.** Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min}$ as in the proof of Proposition \ref{prop:instab}. Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$. If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
\begin{align*}
1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) &= P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}^C) \\
& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
& \le (2^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}
\end{align*}
Then,
\begin{align*}
\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(2^V)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log 2 \rightarrow \infty
\end{align*}
as $N \rightarrow \infty$ by the definition of an unstable DEF model.$\Box$

>>>>

---
title: "A note on the instability and degeneracy of deep learning models"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "October, 2016"
linestretch: 2
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
---

@@

Essentially, in formulating a model for potentially increasing numbers of variables (i.e., $N \rightarrow \infty$), the ratio (\ref{eqn:elpr}) should bounded, meaning that the largest probability under this model at parameter $\boldsymbol \theta_N$ should maintain a magnitude within the smallest probability allowed under the same model. This leads to the following definition:

\begin{defn}[S-unstable DEF]
Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of DEF model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$. A DEF model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), it holds that 
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$.

Note that this definition of *S-unstable* is a generalization or reinterpretation of "unstable" used in @schweinberger2011instability by allowing nonexponential family models and an increasing number of parameters. While this definition appears different, it does match that in @schweinberger2011instability for the special exponential models considered there, see Section \ref{discrete-exponential-family-models} for details.

S-unstable DEF models are undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, 
$$
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*)} : (x_1, \dots, x_N) \& (x_1^*, \dots, x_N^*) \in \mathcal{X}^N \text{ differ by exactly one component}\right\},
$$
the biggest log-probability ratio for a one component change in data outcomes at a DEF parameter $\boldsymbol \theta_N$. We then have the following result.

\begin{prop}
\label{prop:instab}
Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_N)$ be as in (\ref{eqn:elpr}) for an integer $N \ge 1$. If $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$, then $\Delta_N(\boldsymbol \theta_N) > C$.
\end{prop}

In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then the DEF model will exhibit large probability shifts for very small changes in the data configuration (i.e., instability).

**Proof of Proposition \ref{prop:instab}.** We prove the contrapositive. Suppose that $\Delta(\boldsymbol \theta_N) \le C$ holds for some $C > 0$. Under the DEF model, $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ holds for each outcome $\boldsymbol x \in \mathcal{X}^N$. Let $\boldsymbol x_{min} \equiv \argmin\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$ and $\boldsymbol x_{max} \equiv \argmax\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)$. Note there exists a sequence $\boldsymbol x_{min} \equiv \boldsymbol x_0, \boldsymbol x_1, \dots, \boldsymbol x_k \equiv \boldsymbol x_{max}$ in $\mathcal{X}^N$ of component-wise switches to move from $\boldsymbol x_{min}$ to $\boldsymbol x_{max}$ in the sample space (i.e. $\boldsymbol x_i, \boldsymbol x_{i + 1} \in \mathcal{X}^N$ differ by exactly $1$ component for $i = 0, \dots, k$) for some integer $k \in \{0, 1, \dots, N\}$. Then, if $k > 0$, it follows that
\begin{align*}
\log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] &= \left|\sum\limits_{i = 1}^k\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta_N}(\boldsymbol x_{i-1})}\right)\right| \\
& \le \sum\limits_{i = 1}^k\left|\log\left(\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_i)}{P_{\boldsymbol \theta}(\boldsymbol x_{i-1})}\right)\right| \\
& \le k \Delta_N(\boldsymbol \theta_N) \le NC
\end{align*}
using $k \le N$ and $\Delta(\boldsymbol \theta_N) \le C$. If $k = 0$, then $\boldsymbol x_{max} = \boldsymbol x_{min}$ and the same bound above holds.$\Box$

Additionally, S-unstable DEF models are connected to degenerate models, where *degeneracy* implies placing all probability on a small portion of the sample space. Define a modal set 
$$
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{(x_1, \dots, x_N) \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > (1-\epsilon)\max\limits_{(x_1^*, \dots, x_N^*)}P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*) + \epsilon\min\limits_{(x_1^*, \dots, x_N^*)}P_{\boldsymbol \theta_N}(x_1^*, \dots, x_N^*) \right\}
$$
of possible outcomes, for a given $0 < \epsilon < 1$ and along a sequence $\boldsymbol \theta_N$, $N \ge 1$, of DEF parameters.

@@

In other words, as a consequence of S-unstable models, all probability in the DEF model formulation with a large number of random variables will stack up on mode sets or simply those few outcomes with the highest probability.

**Proof of Proposition \ref{prop:degen}.** Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min}$ as in the proof of Proposition \ref{prop:instab}. Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$. If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
\begin{align*}
1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
& \le (2^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}
\end{align*}
Then, it follows that
\begin{align*}
\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(2^V)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log 2 \rightarrow \infty
\end{align*}
as $N \rightarrow \infty$ by the definition of an unstable DEF model. Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$

\begin{remark}
There is a further generalization to the notion of "unstable" used in Schweinberger (2011) that deals with replication. One can conceivably imagine replications of the data $\boldsymbol X_1, \dots, \boldsymbol X_n$, where $X_i \in \mathbb{R}^N, i = 1, \dots, n$ such that there are now $n*N$ random variables in the model. However, the definition of S-unstable and propositions \ref{prop:instab}-\ref{prop:degen} still hold if the replications are independent and identically distributed. The reason for this is that for $n$ observations, $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the model and similarly $\left(\min\limits_{\boldsymbol x \in  \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest. Thus, 
\begin{align*}
\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} &\equiv \frac{1}{n*N}\log \left[\frac{\left\{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right\}^n}{\left\{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)\right\}^n}\right] \\
&= \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right]\\
&= \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N).
\end{align*}
Meaning, the definition of an S-unstable DEF model is invariant to replication.
\end{remark}
